Cover letter for the original IJRR submission

### 2017-10-09 reviews_tro/reviewer_response.tex

\quot{* Finally, I would suggest providing explicit git SHAs or tags for all
of the github links. I've personally made the mistake of linking to a
github project in a paper, only to have that project change and
eventually delete the specific code I was citing. }
\as{The reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project, on a separate branch
dedicated to the installation of the planner.
Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}
\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software. }
\done

\section{Diff files between the original and current submission}
We now join a diff file comparing the original and current submissions.
The text appearing in blue denotes the additions. We used the latexdiff tools to generate this file, which is not fully compatible with all the latex packages that we use.
For this reason, errors appear in the document, in particular regarding the subsections numbering. However, we think that this diff file will help the reviewers locate precisely
where the modifications occured in the document. The reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
are used, and what variables are implicit (that is computed from the others).

\includepdf[pages={-}]{diff.pdf}

### 2017-10-09 reviews_tro/reviewer_response.tex

In the following, we list all the questions and remarks from the reviewers, answer to the questions, and detail the modifications that were brought to the paper.
At the end of this letter, we also join a diff file between the original submission and the present one, that should help the reviewers see the modifications 
brought to the paper.

\section{Answers to reviewer 1}
%~ \ndone reecriture p3 + enlever main video

\quot {As for novel contributions, this did not seem to be addressed/explained
until page 3 (in the last paragraph of Sec. I). Much of the methodology
has already been presented in earlier work, and the particular aspects
that set THIS paper apart seem (as claimed by the authors) to be: 
}

\as{Thanks for the comment. As mentioned in the summary, our planner is almost identical to the one in [20], but the paper comprises important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
for checking static equilibrium.  The main novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20]. This is now explained in details in the introduction.}
\qt{(Section I.B) The present paper is an extension of our ISRR conference paper [20]. 
As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
To validate our contact plans we introduced a complete framework for multi-contact motion synthesis. This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1]. Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28]. 
We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}
\done

@@

\quot{
Provides complete pseudocode.  This is great! However,it's
not fully clear how much of this is novel vs just re-presenting the
same algorithms previously publilshed.	If only code in App. B and C
are new, that should be clear. Conversely, if most of this presentation
was NOT outlined in such detail before, that should also be clear (to
the authors' credit). Either way, a brief clarification in the intro is
recommended.}

\as{We agree with the comment. Indeed, the pseudocode was not present in [20]. We included it in the core of the paper and clarified that it is novel at the end
of Section I.B.}
\qt{Section IV and V of the present paper are thus paraphrasing the original paper, with the exception of Section V.B.2, where we additionally provide the pseudo-code of the algorithm absent from the original paper.}
\done

\quot{(3)	The contribution to provide a solution to P3 doesn't happen
until page 13 (Appendix C) and as with contribution (1), it needs a
better, intuitive summary (briefly) when first mentioned, as well as a
roadmap for the reader as to where it will be described. It is also
ambiguous in the paper whether to what extent (if any) this solution is
novel, versus a presentation of work from [1]. "Provide a solution"
seemed to imply novelty, in the context of this paragraph -- yet later,
the authors (near the end of page 2) state they use a "state-of-the-art
solution to P3", which implies the P3 solution is not a claim of
novelty.}

\as{The main contribution of the paper is the proposal and validation of an algorithm for P1 and P2. 
In the new version of the paper, we do not consider that our solution to P3
is a new contribution, although our formulation differs from [1] (our problem variables are the accelerations
of the center of mass, whereas in [1] the problem variables are the contact forces). We provide it here for completeness, because it corresponds to our implementation, and 
was required to validate our approach. 
We rephrased the text to clearly indicate that the algorithm is not a contribution. We made the choice not to give an intuition nor describe the solver in the core
of the paper because it is not required to understand the contributions to P1 and P2.}
\qt{(Section I.B) To validate our contact plans we introduced a complete framework for multi-contact motion synthesis. This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1]. Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28]. 
We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.}\done

@@

\section{Answers to reviewer 6}
\quot{ My primary concern, however,
is that the elements described in this paper have been described in
previous published work, which makes the particular contributions of
this work unclear. To be specific, the reachability-guided planner
appears identical to the planner in [20] (as stated by the authors in
the end of Section I.B.), and the optimal control formulation is from [1]
(again as stated in I.B.).}
\as{Thanks for the comment. Indeed, all reviewers agreed that the contribution with respect to previous work was not clearly stated. We updated the paper to reflect the following. Firstly, our planner is almost identical to the one in [20], with important additions that are the pseudo code for the algorithm (the algorithm is not given in [20]), as well as the introduction of a more efficient criterion
for checking static equilibrium (see next question). The optimal control is no longer a claimed contribution of the paper, even though it differs from [1] in its formulation. The main
novelty of the paper resides in the experimental setup and analysis, as well as the empirical proof that the planner works with real robots that are more constrained kinematically than the virtual avatars from [20].}
\qt{(Section I.B) 
The present paper is an extension of our ISRR conference paper [20]. 
As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section IV and V). 
However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section V.B.2), a novel criterion for
static equilibrium (Section VI), and the release of the source code of the planner (Section VII). \\
The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section VIII).
To validate our contact plans we introduced a complete framework for multi-contact motion synthesis. This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work [1]. Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm [28]. 
We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution. \\
Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.}\done

### 2017-10-09 body/intro.tex
Review response to reviewer
rewrote I.B more like in the letter

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling has been shown by Escande et al. \cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. Bouyarmane et al. ~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path. Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment. This planning phase does not account for collisions, implying that re-planning is required in case of failure. In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}. A limitation with
these approaches (including our method) is that the existing planners only address a subset of the problem, because their ability to find a solution depends
on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case. Other contributions to legged locomotion, not directly related to multi-contact motion, are worth mentioning, as they rely on a similar decomposition of the problem. First a discrete sequence 
of contact sets is planned (at the so-called footstep planning phase), using a low-dimensional abstraction of the robot to account for its kinematic constraints~\cite{6078435, doi:10.1177/0278364910392608}. In \cite{doi:10.1177/0278364910392608}, a ``pose certificate'' is obtained by generating a whole-body configuration for each set through inverse kinematics, as done in $\mathcal{P}_2$. Then a
motion is generated along the sequence through the use of optimization techniques. 
The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks. They thus cannot be generalized to to multi contact locomotion. 
%For instance, the choice of the next effector in contact is often deterministic, the contacts are either coplanar or punctual, or the generation of the end-effector trajectory only works for ``ground'' obstacles, as it assumes that increasing the $z$ position of an effector drives it further away from obstacles~\cite{Kolter:2009:TTV:1703775.1703833}. 
However, some contributions on specific parts of the problem could be applied directly in our case. 
For instance, learning ``terrain costs'', based on expert knowledge as proposed in ~\cite{Ratliff:2009:LSF:1569248.1569253}, could define good heuristics to compute the next contact location for an effector. Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.

@@

\noindent \textbf{Comparison with our previous work:}
The present paper is an extension of our conference paper~\cite{tonneauisrr15}. As such, the motion planning algorithm presented here is the same as the one presented in the
previous work. Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper. We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section~\ref{sec:Equil}.
Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
all the technical information required to use the planner.

To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team. We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work~\cite{Carpentier2016}. The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework. Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.

Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
The novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are much more constrained kinematically than the virtual avatars used before,
thus making the planning problem even harder. Through an extensive empirical analysis, we demonstrate that the method remains interesting in this context, qualitatively and efficiently.

%~ This paper extends our previous conference paper~\citep{tonneauisrr15} in several ways. Our approach is now demonstrated on actual robots and in simulation, contrary to virtual avatars before, in a large variety of contexts. To achieve this we introduce a new criterion to efficiently compute the equilibrium of the robot while accounting for uncertainties of the real world. We provide the complete pseudo code of algorithm. We give the complete procedure to use our planner with any legged robot and provide our source code, allowing any user to directly use our method. We also propose a solution to $\mathcal{P}_3$, making our framework complete. Finally we provide an in-depth analysis of the results we obtain, and a theoretical discussion on the capabilities of the planner and its current limitations.

>>>>

\textbf{On the other hand, the two problems \Pa and \Pb might be decoupled} to reduce the complexity. The feasibility and interest of the decoupling has been shown by Escande et al. \cite{DBLP:conf/iser/EscandeKMG08} who manually set up a rough root guide path (i.e. an ad-hoc solution to $\mathcal{P}_1$), and then addressed \Pb as the combinatorial computation of a feasible contact sequence in the neighborhood of the guide. A solution could then be found, %efficiently when considering quadruped locomotion \citep{kalakrishnan2011learning}, 
but at the cost of prohibitive computation times (up to several hours) for constraining scenarios. This approach is suboptimal because the quality of the motion depends on the quality of the guide path. Bouyarmane et al. ~\cite{Bouyarmane2009} precisely focused on automatically computing a guide path with guarantees of \textit{equilibrium feasibility}, by extending key frames of the path into whole-body configurations in static equilibrium. Randomly-sampled configurations are projected to the contact manifold using an inverse-kinematics solver, a computationally-expensive process (about 15 minutes to compute a guide path in the examples presented). Moreover this explicit projection is insufficient to guarantee the feasibility between two key postures in the path. Chung and Khatib~\cite{7140082} also proposed a decoupled approach, with a planning phase based on the reachable workspace of the robot limbs, used to judge the ability to make contact with a discretized environment. This planning phase does not account for collisions, implying that re-planning is required in case of failure. In highly-constraining cases such as the car egress scenario we address, we believe that including collision constraints in the planning is a requirement~\citep{tonneauisrr15,grey2017footstep}. A limitation with
these approaches (including our method) is that the existing planners only address a subset of the problem, because their ability to find a solution depends
on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case. Other contributions to legged locomotion, not directly related to multi-contact motion, are worth mentioning, as they rely on a similar decomposition of the problem. First a discrete sequence 
of contact sets is planned (at the so-called footstep planning phase), using a low-dimensional abstraction of the robot to account for its kinematic constraints~\cite{6078435, doi:10.1177/0278364910392608}. In \cite{doi:10.1177/0278364910392608}, a ``pose certificate'' is obtained by generating a whole-body configuration for each set through inverse kinematics, as done in $\mathcal{P}_2$. Then a
motion is generated along the sequence through the use of optimization techniques. 
The solutions proposed are designed for cyclic locomotion in quasi-flat scenarios, where the support polygon is a relevant method for equilibrium checks. They thus cannot be generalized to multi contact locomotion. 
%For instance, the choice of the next effector in contact is often deterministic, the contacts are either coplanar or punctual, or the generation of the end-effector trajectory only works for ``ground'' obstacles, as it assumes that increasing the $z$ position of an effector drives it further away from obstacles~\cite{Kolter:2009:TTV:1703775.1703833}. 
However, some contributions on specific parts of the problem could be applied directly in our case. 
For instance, learning ``terrain costs'', based on expert knowledge as proposed in ~\cite{Ratliff:2009:LSF:1569248.1569253}, could define good heuristics to compute the next contact location for an effector. Although we did not try to include such formulation in this paper, it would be straightforward to integrate such heuristic in our planner.

@@

\noindent \textbf{Comparison with our previous work:}
The present paper is an extension of our ISRR conference paper~\cite{tonneauisrr15}. 
As such, our solution to address $\mathcal{P}_1$ and $\mathcal{P}_2$ is the same motion planner as the one presented at ISRR (reformulated in Section~\ref{rbprm} and~\ref{sec:contact}). 
However three important novelties have been added to the planner: the pseudo-code of the algorithm (Section~\ref{app:contact}), a novel criterion for
static equilibrium (Section~\ref{sec:Equil}), and the release of the source code of the planner (Section~\ref{app:hpp}).

The other contribution of the paper is a rigorous experimental validation of the approach on actual robot models (Section~\ref{sec:results}).
To validate our contact plans we introduced a complete framework for multi-contact motion synthesis. This framework additionally comprises an interpolation method to solve the problem $\mathcal{P}_3$,  based on a reformulation of our previous work~\cite{Carpentier2016}. Our solution to $\mathcal{P}_3$ allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}. 
We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand our experimental contribution.

Finally, as opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method remains efficient and applicable in this context.
%~ These aspects of the framework are presented in details in the paper, but are not novel \textit{per se}. The novelty lies in the validation of the method with real robot models, for which the planning is harder with respect to the avatars used in [20], because of stricter kinematic constraints

%~ As such, the motion planning algorithm presented here is the same as the one presented in the
%~ previous work. Section~\ref{rbprm} and~\ref{sec:contact} of the present paper are thus paraphrasing the original paper, with the exception of Section~\ref{app:contact}, where we additionally provide the pseudo-code of the algorithm absent from the original paper. We also improve the computational performance of the method by replacing the static equilibrium test with a more efficient test based on a new LP formulation presented in Section~\ref{sec:Equil}.
%~ Compared with the original paper, we also release our source code, and provide the complete procedure to use our planner for any legged robot. We thus provide
%~ all the technical information required to use the planner.
%~ 
%~ To validate our approach, in this paper we had to implement or integrate other tools based on previous publications from our team. We implemented a solution to $\mathcal{P}_3$ based on a reformulation of our previous work~\cite{Carpentier2016}. The implementation of this solution is also provided as an open-source project, which can be integrated directly in our framework. Our solution to $\mathcal{P}_3$ then allows us to verify that the synthesized motions are physically consistent, using our implementation of a state-of-the-art simulation algorithm~\cite{Kaufman2008}.
%~ We do not consider these additions as contributions \textit{per se}, but describing them here is required to understand the experimental contribution of this paper.

%~ Indeed, thanks to these tools we have been able to design a complete pipeline for synthesizing multi-contact motions from scratch.
%~ Indeed, the novelty of the present paper lies in the rigorous experimental setup we propose to demonstrate its interest.
%~ As opposed to our previous paper, we demonstrate the algorithms with actual robotic models. These models are more constrained kinematically than the virtual avatars used before,
%~ thus making the planning problem harder. Through an extensive empirical analysis, we demonstrate that the method is remains efficient and qualitatively relevant in this context.
